Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2015-01-12 response_to_reviewers.tex

\textbf{16. Reviewer's Comment:}
Page 7, final sentence of results: The test of Apocynaceae feeding is not presented in the Methods section.

\textbf{16. Response:}

\begin{quote}
\color{blue}
We added the following sentence to the Methods section:

We also run a BiSSE analysis
to test whether the trait ``feeding on Apocynaceae'' had any effect on
net diversification rates found for Nymphalidae lineages feeding on
Apocynaceae.
\end{quote}

\textbf{17. Reviewer's Comment:}
Page 7, first and third paragraphs of Discussion section (and potentially elsewhere): there are cases where the term “diversification splits” is used, where “diversification shifts” is more appropriate.

\textbf{17. Response:}

\begin{quote}
\color{blue}
We changed ``splits'' to ``shifts'' throughout the text.
\end{quote}

\textbf{18. Reviewer's Comment:}
Page 7, sentence beginning “For example, in our Nymphalidae trees...”: the authors state “we found that the split...had a variation of r = 0.5234”. Should “variation” instead be “rate”?

\textbf{18. Response:}

\begin{quote}
\color{blue}
\end{quote}

\end{document}

### 2015-01-12 MS_revised.tex
comment 17

It is also evident that not all the diversification shifts estimated on
the MCC tree are consistently recovered in most of the 1000 trees. Some
of the shifts in the MCC tree are recovered in very few trees, for
example the split for the clade Satyrinae is recovered with a
probability of 0.18 (Fig. 3).

@@

The MEDUSA method has been used to infer changes in net diversification
rates in a phylogenetic tree. Since its publication \cite{alfaro2009} the results
of using MEDUSA on a single tree, the maximum clade credibility tree,
have been used for generation of hypotheses and discussion
\cite{heikkila2012, litman2011, ryberg2012}. However, different diversification shifts and different
rates of diversification are found for certain lineages when
phylogenetic uncertainty was taken into account by using MEDUSA on a
random sample of trees from the posterior distribution of a Bayesian
run. We found that some diversification shifts, estimated on the
Nymphalidae maximum clade credibility tree, were found with very low
probability in the random sample of 1000 trees from the posterior
distribution (Fig. 3, Table 2). We also found that, even though the
analyses estimated the same diversification shifts on two or more trees,
the estimated net diversification rates could vary widely (Fig. 2). For
example, in our Nymphalidae trees, we found that the split for
\emph{Magneuptychia} and \emph{Caeruleuptychia} had a variation from
\texttt{r = 0.5234}, higher than the background net diversification
rate, to almost zero. This means that observed patterns and conclusions
can be completely contradictory depending on tree choice.

@@

% Table 1
\section*{Tables}
\begin{table}[!h]
\caption{\bf{Significant net diversification rate shifts found in the MEDUSA analysis of the Nymphalid maximum clade credibility tree.}}
\begin{tabular}{lccrrcl}
Shift N$^\circ$ & Split.Node & Model & r          & LnLik.part & AICc     & Taxa                                                       \\
1               & 399        & yule  & 0.092459   & -1055.957  &          & Nymphalidae (root)                                         \\
2               & 691        & bd    & 0.054129   & -406.3703  &          & Limenitidinae + Heliconiinae                               \\
3               & 299        & yule  & 0.311199   & -6.3058    &          & \emph{Ypthima}                                             \\
4               & 224        & yule  & 0.290989   & -6.2601    &          & \emph{Charaxes}                                            \\
5               & 750        & yule  & 0.186913   & -147.4146  &          & Oleriina + Ithomiina + Napeogenina + Dircennina + Godyrina \\
6               & 405        & yule  & 0.116252   & -555.0276  &          & Satyrinae                                                   \\
7               & 495        & yule  & 0.064656   & -124.9143  &          & Coenonymphina                                              \\
8               & 609        & yule  & 0.240562   & -35.0908   &          & Phyciodina in part                                         \\
9               & 787        & bd    & 0.042332   & -43.7819   &          & Danaini in part                                            \\
10              & 231        & yule  & 0.209416   & -4.7955    &          & \emph{Coenonympha}                                         \\
11              & 478        & yule  & 0.311684   & -9.2218    &          & \emph{Caeruleuptychia} + \emph{Magneuptychia}              \\
12              & 659        & yule  & 0.219253   & -11.2686   &          & \emph{Callicore} + \emph{Diaethria}                        \\
13              & 444        & yule  & 0.220615   & -43.7812   &          & Satyrina                                                   \\
14              & 524        & yule  & 0.190754   & -26.0651   &          & Mycalesina                                                 \\
15              & 355        & yule  & 0.234041   & -6.2013    &          & \emph{Pedaliodes}                                          \\
16              & 714        & yule  & 0          & 0          &          & \emph{Dryas} + \emph{Dryadula}                             \\
17              & 377        & yule  & 0.311671   & -4.1986    &          & \emph{Taenaris}                                            \\
18              & 688        & yule  & 0.024724   & -17.5256   &          & Pseudergolinae                                             \\
19              & 583        & yule  & 0          & 0          & 5090.492 & \emph{Anaeomorpha} + \emph{Hypna}                                       
\end{tabular}
\begin{flushleft}Split.Node=node number, Model=preferred diversification model by MEDUSA, r=net diversification rate, LnLik.part=log likelihood value.
\end{flushleft}
\end{table}

% Table 2
\begin{table}[!h]
    \caption{\bf{Differences in rates estimated by MEDUSA on the MCC
tree from the sample of trees from the posterior distribution and the
MultiMEDUSA approach. Shift consistently recovered across the
sample of trees in bold face.}}
\begin{tabular}{lcrr}
Split.Node & rate by MEDUSA & Median rate by MultiMEDUSA & probability of being recovered \\
1          & 0.092          & not found                  & 0.000                          \\
2          & 0.055          & -0.030                     & 0.864                          \\
\bf{3}     & 0.184          & \bf{0.219}                 & \bf{0.961}                     \\
\bf{4}     & 0.166          & \bf{0.212}                 & \bf{0.996}                     \\
\bf{5}     & 0.111          & \bf{0.101}                 & \bf{0.927}                     \\
6          & 0.119          & 0.039                      & 0.131                          \\
7          & 0.066          & -0.052                     & 0.195                          \\
8          & 0.232          & 0.166                      & 0.319                          \\
9          & 0.042          & -0.049                     & 0.897                          \\
10         & 0.058          & 0.149                      & 0.619                          \\
11         & 0.311          & 0.208                      & 0.831                          \\
\bf{12}    & 0.219          & \bf{0.135}                 & \bf{0.911}                     \\
13         & 0.082          & 0.117                      & 0.276                          \\
14         & 0.099          & 0.115                      & 0.379                          \\
15         & 0.113          & 0.127                      & 0.825                          \\
16         & 0.222          & -0.005                     & 0.659                          \\
17         & 0.243          & 0.248                      & 0.785                          \\
18         & 0.192          & -0.064                     & 0.024                          \\
19         & 0.064          & -0.087                     & 0.381                         
\end{tabular}
\end{table}

>>>>

It is also evident that not all the diversification shifts estimated on
the MCC tree are consistently recovered in most of the 1000 trees. Some
of the shifts in the MCC tree are recovered in very few trees, for
example the shift for the clade Satyrinae is recovered with a
probability of 0.18 (Fig. 3).

@@

The MEDUSA method has been used to infer changes in net diversification
rates in a phylogenetic tree. Since its publication \cite{alfaro2009} the results
of using MEDUSA on a single tree, the maximum clade credibility tree,
have been used for generation of hypotheses and discussion
\cite{heikkila2012, litman2011, ryberg2012}. However, different diversification shifts and different
rates of diversification are found for certain lineages when
phylogenetic uncertainty was taken into account by using MEDUSA on a
random sample of trees from the posterior distribution of a Bayesian
run. We found that some diversification shifts, estimated on the
Nymphalidae maximum clade credibility tree, were found with very low
probability in the random sample of 1000 trees from the posterior
distribution (Fig. 3, Table 2). We also found that, even though the
analyses estimated the same diversification shifts on two or more trees,
the estimated net diversification rates could vary widely (Fig. 2). For
example, in our Nymphalidae trees, we found that the shift for
\emph{Magneuptychia} and \emph{Caeruleuptychia} had a variation from
\texttt{r = 0.5234}, higher than the background net diversification
rate, to almost zero. This means that observed patterns and conclusions
can be completely contradictory depending on tree choice.

@@

% Table 1
\section*{Tables}
\begin{table}[!h]
\caption{\bf{Significant net diversification rate shifts found in the MEDUSA analysis of the Nymphalid maximum clade credibility tree.}}
\begin{tabular}{lccrrcl}
Shift N$^\circ$ & Shift.Node & Model & r          & LnLik.part & AICc     & Taxa                                                       \\
1               & 399        & yule  & 0.092459   & -1055.957  &          & Nymphalidae (root)                                         \\
2               & 691        & bd    & 0.054129   & -406.3703  &          & Limenitidinae + Heliconiinae                               \\
3               & 299        & yule  & 0.311199   & -6.3058    &          & \emph{Ypthima}                                             \\
4               & 224        & yule  & 0.290989   & -6.2601    &          & \emph{Charaxes}                                            \\
5               & 750        & yule  & 0.186913   & -147.4146  &          & Oleriina + Ithomiina + Napeogenina + Dircennina + Godyrina \\
6               & 405        & yule  & 0.116252   & -555.0276  &          & Satyrinae                                                   \\
7               & 495        & yule  & 0.064656   & -124.9143  &          & Coenonymphina                                              \\
8               & 609        & yule  & 0.240562   & -35.0908   &          & Phyciodina in part                                         \\
9               & 787        & bd    & 0.042332   & -43.7819   &          & Danaini in part                                            \\
10              & 231        & yule  & 0.209416   & -4.7955    &          & \emph{Coenonympha}                                         \\
11              & 478        & yule  & 0.311684   & -9.2218    &          & \emph{Caeruleuptychia} + \emph{Magneuptychia}              \\
12              & 659        & yule  & 0.219253   & -11.2686   &          & \emph{Callicore} + \emph{Diaethria}                        \\
13              & 444        & yule  & 0.220615   & -43.7812   &          & Satyrina                                                   \\
14              & 524        & yule  & 0.190754   & -26.0651   &          & Mycalesina                                                 \\
15              & 355        & yule  & 0.234041   & -6.2013    &          & \emph{Pedaliodes}                                          \\
16              & 714        & yule  & 0          & 0          &          & \emph{Dryas} + \emph{Dryadula}                             \\
17              & 377        & yule  & 0.311671   & -4.1986    &          & \emph{Taenaris}                                            \\
18              & 688        & yule  & 0.024724   & -17.5256   &          & Pseudergolinae                                             \\
19              & 583        & yule  & 0          & 0          & 5090.492 & \emph{Anaeomorpha} + \emph{Hypna}                                       
\end{tabular}
\begin{flushleft}Shift.Node=node number, Model=preferred diversification model by MEDUSA, r=net diversification rate, LnLik.part=log likelihood value.
\end{flushleft}
\end{table}

% Table 2
\begin{table}[!h]
    \caption{\bf{Differences in rates estimated by MEDUSA on the MCC
tree from the sample of trees from the posterior distribution and the
MultiMEDUSA approach. Shift consistently recovered across the
sample of trees in bold face.}}
\begin{tabular}{lcrr}
Shift.Node & rate by MEDUSA & Median rate by MultiMEDUSA & probability of being recovered \\
1          & 0.092          & not found                  & 0.000                          \\
2          & 0.055          & -0.030                     & 0.864                          \\
\bf{3}     & 0.184          & \bf{0.219}                 & \bf{0.961}                     \\
\bf{4}     & 0.166          & \bf{0.212}                 & \bf{0.996}                     \\
\bf{5}     & 0.111          & \bf{0.101}                 & \bf{0.927}                     \\
6          & 0.119          & 0.039                      & 0.131                          \\
7          & 0.066          & -0.052                     & 0.195                          \\
8          & 0.232          & 0.166                      & 0.319                          \\
9          & 0.042          & -0.049                     & 0.897                          \\
10         & 0.058          & 0.149                      & 0.619                          \\
11         & 0.311          & 0.208                      & 0.831                          \\
\bf{12}    & 0.219          & \bf{0.135}                 & \bf{0.911}                     \\
13         & 0.082          & 0.117                      & 0.276                          \\
14         & 0.099          & 0.115                      & 0.379                          \\
15         & 0.113          & 0.127                      & 0.825                          \\
16         & 0.222          & -0.005                     & 0.659                          \\
17         & 0.243          & 0.248                      & 0.785                          \\
18         & 0.192          & -0.064                     & 0.024                          \\
19         & 0.064          & -0.087                     & 0.381                         
\end{tabular}
\end{table}
